### 2017-05-12 lafaySimulatedScenePerception.tex
ml spell check done

Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement. Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has. Unfortunately, without any reference to sound data, this representation cannot be caracterized physically.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify the representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is saked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both caracterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

Recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}. For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very insteresting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest. Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characterestics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmanelta scenes. Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound soucres typically present in a urban scene impact pleasantness:

@@

A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire. Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part. Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établi. Les auteurs montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques. Ils soulignent néanmoins que le manque d'enregistrements disponibles limite l'analyse.

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical caracteristics. A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added befenit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{SimScene} is introduced in Section \ref{sec:simulator}. The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modiciation of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

\emph{Simscene} is designed to be used as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sounds source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

Litterature usually distinguish between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely ditrubuted whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and consider two disctint sound databases: one with the classes of events only and the other with classes of textures. These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.

@@

We do so, because we assume that the morphological differences between those two classes have some important consequences on the percpetion of the scenes.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorisation processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".

@@

We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples percieved equivalently.

@@

Orignal sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant eareas with no dominant sound sources.

@@

All samples are normalised to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.

@@

By design, the simulation tool do not allow the user to interect and control a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several paramters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for the event's track only, the spacings arerandomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject set the first and last occurence of a sample of the selected class.
\end{itemize}

@@

The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available ofr this kind of track.

@@

The previous experiment demonstrated that, among the classes of sounds occuring in urban soundscapes, those with markers are specific of some of environments. Those sound markers seem to have a great impact on perception. This impact is studied in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.

@@

In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveforms of the simulated scenes are regenerated without considering the classes identified as markers. For the i-scenes, the i-markers are removed. For the ni-scenes, the ni-markers are removed. A perceptive test evaluating the pleasantness is conducted with a protocol close to the one considered in experiment 1.b.

@@

Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the percieved pleasantness. Two hypothesis are formulated:

@@

\subsection{Planification of experiment 2}

@@

There are 144 stimuli of 30 seconds durations. More precisely:

\begin{itemize}
\item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept. The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
\item \emph{72 rm-scènes}:  72 scenes where the sound classes identified as markers are removed. The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
\end{itemize}

@@

Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactely the same.

@@

\begin{itemize}
\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} pour les i/sm-scènes;
\item \emph{siren}, \emph{car horn} pour les ni/sm-scènes.
\end{itemize}

@@

Les sujets évaluent les 144 scènes. L'évaluation s'effectue sur une échelle sémantique bipolaire de 11 points allant de -5 (non-idéale/très désagréable) à +5 (idéale/très agréable). Avant de noter une scène, les sujets doivent obligatoirement en écouter les 20 premières secondes. Après la notation, ils sont libres de passer à la scène suivante.

Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire. Les 10 premières scènes permettent au sujet de calibrer ses notes. Elles sont obligatoirement composées de 5 i/am-scènes et de 5 ni/am-scènes. Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.

L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.

@@

12 subjects (4 womens) perform the test. None of them have performed experiments 1.a and 1.b. The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14). All the subjects live in Nantes, France, for at least two years. All the subjects performed the test succesfully.

@@

The type of data analysed in this experiment have been considered for experiment 1.a, see Section~\ref{sec:xp1_dataAna} for details.

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a variance analysis. We consider $\mathcal{A}_{sujet}$ as a dependant variable, and as independant variables, the type of envorinment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of percieved pleasantness between km-scenes and rm-scenes. The two independant variables are cosidered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity impothesis does not need to be checked. \emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.

@@

In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes. The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/abscence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the percieved pleasantness. The removal of the ni-markers has a positive effect on the perception of the ni-scenes. Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes. This is even more striking because, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.

@@

This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene et a non ideal one.

@@

That being said, we observe that the physical caracteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes. In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that proeminently influences the perceived quality.

@@

These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into accound during the perceptual process appearing to vary from one source to the other, from one type of environment to the other. This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.

@@

\begin{itemize}
\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, wether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
\item in the case of ideal scenes, one should first idenity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
\end{itemize}

@@

Third, the fact that the compositional descriptors correlated with the percieved pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurences in the simulated scene being directly available.
\end{itemize}

>>>>

%Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement. Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has. Unfortunately, without any reference to sound data, this representation cannot be characterized physically.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify the representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

Recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}. For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest. Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

%A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire. Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part. Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établi. Les auteurs montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques. Ils soulignent néanmoins que le manque d'enregistrements disponibles limite l'analyse.

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

\emph{Simscene} is designed to be used as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sounds source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

Literature usually distinguish between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures. These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.

@@

We do so, because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".

@@

We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant areas with no dominant sound sources.

@@

All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.

@@

By design, the simulation tool do not allow the user to interact and control a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for the event's track only, the spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
\end{itemize}

@@

The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available for this kind of track.

@@

The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific of some of environments. Those sound markers seem to have a great impact on perception. This impact is studied in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.

@@

In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated without considering the classes identified as markers. For the i-scenes, the i-markers are removed. For the ni-scenes, the ni-markers are removed. A perceptive test evaluating the pleasantness is conducted with a protocol close to the one considered in experiment 1.b.

@@

Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are formulated:

@@

\subsection{Planning of experiment 2}

@@

There are 144 stimuli of 30 seconds duration. More precisely:

\begin{itemize}
\item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept. The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
\item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed. The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
\end{itemize}

@@

Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactly the same.

@@

\begin{itemize}
\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} for the i/sm-scenes;
\item \emph{siren}, \emph{car horn} for the ni/sm-scenes.
\end{itemize}

@@

%Les sujets évaluent les 144 scènes. L'évaluation s'effectue sur une échelle sémantique bipolaire de 11 points allant de -5 (non-idéale/très désagréable) à +5 (idéale/très agréable). Avant de noter une scène, les sujets doivent obligatoirement en écouter les 20 premières secondes. Après la notation, ils sont libres de passer à la scène suivante.

%Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire. Les 10 premières scènes permettent au sujet de calibrer ses notes. Elles sont obligatoirement composées de 5 i/am-scènes et de 5 ni/am-scènes. Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.

%L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.

The subjects evaluate the 144 scenes. The evaluation is done on an 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant). Before grading a scene, subjects must listen to the first 20 seconds. After scoring, they are free to move on to the next scene.

For each subject, the scenes are presented in random order. The first 10 scenes allow the subject to calibrate his/her grades. Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes. These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.

The experiment is scheduled to last 1 hour. The subjects do not know the nature of the scenes.

@@

12 subjects (4 women) perform the test. None of them have performed experiments 1.a and 1.b. The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14). All the subjects live in Nantes, France, for at least two years. All the subjects performed the test successfully.

@@

The type of data analyzed in this experiment have been considered for experiment 1.a, see Section~\ref{sec:xp1_dataAna} for details.

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a variance analysis. We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.

@@

In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes. The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness. The removal of the ni-markers has a positive effect on the perception of the ni-scenes. Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes. This is even more striking because, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.

@@

This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene and a non ideal one.

@@

That being said, we observe that the physical characteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes. In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that prominently influences the perceived quality.

@@

These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into account during the perceptual process appearing to vary from one source to the other, from one type of environment to the other. This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.

@@

\begin{itemize}
\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
\end{itemize}

@@

Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
\end{itemize}
